---
title: "Exercise 4: Reproduction number"
author: Éric Marty
date: "May 24, 2022"
header-includes:
   - \usepackage{array}
   - \tikzstyle{custom}=[fill=white, draw=black]
output:
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: ../modelgraph.tex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../R/modelgraph.R")
```

Complete Exercise 4a. If you have time, look at exercise 3b.

## Exercise 4a

Following is the state diagram for an SEIR model with balanced birth and death. $\mu$ is the birth *and* death rate, implying zero net population change. $\beta$ is the transmission rate, $\sigma$ is the progression rate from exposed to infected, and $\gamma$ is the recovery rate. 

```{r SEIR, echo = FALSE}
states <- data.frame(
  name = c(c("S", "E", "I", "R"),
           c("in","Sout", "Eout", "Iout", "Rout")
           ),
  label = c(c("$S$", "$E$", "$I$", "$R$"),
            rep("",5)
            ),
  class = c(c("susceptible", "exposed", "infectious", "recovered"),
            rep("invisible",5)
            ),
  position = c("",paste0("right of=",c("S", "E", "I")),
               c("above of=S",paste0("below of=",c("S", "E", "I", "R")))
               )
)

flows <- data.frame(
  from = c(c("S", "E", "I"),
           c("in","S", "E", "I", "R")
           ),
  to = c(c("E", "I", "R"),
           c("S", "Sout", "Eout", "Iout", "Rout")
           ),
  label = c(c("$\\beta S I$","$\\sigma E$", "$\\gamma I$"),
           c("$\\mu (S+E+I+R)$", "$\\mu S$", "$\\mu E$", "$\\mu I$", "$\\mu R$")
           ),
  label.properties = c(c("above", "above", "above"),
           c("right", "right", "right", "right", "right")
           )
)

# Compile a snipet of LaTeX from the dataframes of states and flows
SEIR <- modelgraph(states, flows)
```

\resizebox{0.35\textwidth}{!}{`r SEIR`}

$\mathcal{R}_0$ for this model can be estimated as: 


$$\mathcal{R}_0 = \frac{\beta\sigma}{(\mu+\gamma)(\mu+\sigma)},$$

where $\mu \ll \gamma$ and $\sigma$ is of the order of $\gamma$. 

Your task is to design a plot to explore variation in $\mathcal{R}_0$ across and range of $\gamma$ and $\sigma$. 

You should complete the exercise in a new Rmarkdown document.

### Step 1: 

Begin by fixing $\mu$ and $\beta$ and preparing a grid of $\gamma$ and $\sigma$. (You can use `expand.grid()` for this.) Consider the question(s) are you trying to answer about the model. How fine of a grid you will need? What should the bounds be?

Calculate $\mathcal{R}_0$ for all values of the grid.

Transform your data into a matrix of $\mathcal{R}_0$ values, 
with row indices = levels of $\gamma$, and column indices = levels of $\mu$.

\color{blue}
*To see a solution, open the Rmd version of this file ("exercises/4_reproduction_number.Rmd") and look at the hidden code chunk at this point in the document.*
\color{black}

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
# Example with fixed beta, fixed mu, and grid of gamma and sigma
beta <- .15
mu <- .01
gamma <- seq(1/10, 1/3,length.out=20)
sigma <- seq(1/10,1,length.out=20)
results <- expand.grid(gamma = gamma, sigma = sigma)

# Calculate R_0
# Note: In tidyverse, use all_of(varname) to refer to an object outside the dataframe.
# Use varname to refer to column of the dataframe.

results <- results %>% 
  mutate(R0 = all_of(beta) * sigma / ((all_of(mu) + gamma)*(all_of(mu) + sigma)))

# Prepare a matrix
results.wide <- results %>% 
  pivot_wider(names_from = sigma, values_from = R0)
results.matrix <- results.wide %>% 
  select(c(-gamma)) %>% 
  as.matrix()
dimnames(results.matrix) <- list(gamma = results.wide$gamma,
                                 sigma = colnames(results.matrix))

```

### Step 2: 

Visualize.

A matrix of smoothly changing values can be visualized as a "level plot" 
(with values mapped to colors), or as a "contour plot" 
(with contours drawn at set intervals), or both.

A level plot can be made with the `lattice::levelplot()` or `graphics::filled.contour()`. 
Both functions can add contours to the level plot.

**Questions to consider:**

How might aspects of the visual design help reveal answers to questions you have about the model?

What kind of color scale would be appropriate to represent $\mathcal{R}_0$? (Think carefully about what $\mathcal{R}_0$ represents, the critical value(s), boundaries, etc.)

Does a log scale make sense for the independent variables?

Does a log scale make sense for the dependent variable ( $\mathcal{R}_0$)? 
If so, how would you construct a log color scale?

Here is a default plot of a matrix of $\mathcal{R}_0$ by $\sigma$ and $\gamma$ using `filled.contour()`:

```{r}
filled.contour(results.matrix)
```

Problems with the default plot:

1. Axes are not labeled and do not display the actual parameter values.
2. It is difficult to read data off the plot.
3. The color scale does not have an obvious semantic relationship to $\mathcal{R}_0$.  It is not immediately apparent where the critical value of $\mathcal{R}_0$ lies. It is not immediately intuitive what the colors mean. (Could you guess at the $R_0$ values without the legend?)

\color{blue}
*To see a solution, open the Rmd version of this file ("exercises/4_reproduction_number.Rmd") and look at the hidden code chunk at this point in the document.*
\color{black}

```{r, echo=FALSE, eval=FALSE}
# 1. Prepare an appropriate color scale...

# In this case, we have chosen a divergent color scale to be centered around R_0 = 1.
# In addition, we will map color to the log of R_0, 
# since there are no values of R_0 < 0, and we'd like to have an equally 
# wide range of colors above and below R_0 = 1. 

# define color palette function
colorfunction <- colorRampPalette(
  c("blue", "white", "red"), # start, middle and end colors
  space = "Lab"  # specifies a perceptually linear color space (LAB)
  ) # returns a function taking 1 argument (ncolors)

# Set number of levels in the the color scale
# Choose an even number to have an equal number of colors above and below R_0 = 1.
nlevels <- 20

# Build the actual color palette:
colors <- colorfunction(nlevels) # one fewer colors than breaks

# Preview the colors as a color bar:
scales::show_col(colors, ncol = length(colors), labels = FALSE)
```

```{r, echo=FALSE, eval=FALSE}

# define an appropriate range encompassing all values of log(R_0)
logR0_lowerlim <- floor(min(log(results.matrix)))
logR0_upperlim <- ceiling(max(log(results.matrix)))
logR0_abslimit <- max(abs(logR0_lowerlim),abs(logR0_upperlim))

# Set the break points for R_0 in the color scale (i.e. the contours).
# We need one more break point than levels in the color scale (nlevels + 1).
# Using ±logR0_abslimit for the endpoints will ensure a symetrical log scale 
# encompassing all values of log(R_0). 
# The sequence must then be retransformed to get the breakpoints in natural units of R_0.
breaks <- exp(seq(-logR0_abslimit, logR0_abslimit, length.out = nlevels+1))

# Plot
filled.contour(results.matrix,
               levels = breaks, # levels of R_0 defining contours
               col = colors # vector of colors (n = 1 - breaks)
               )
```

```{r, echo=FALSE, eval=FALSE}
# 2. Define and label axes, fix up legend

xtick.label <- format(as.numeric(colnames(results.matrix)),digits=2)
xtick.pos <- seq(0,1,length.out = length(xtick.label))

ytick.label <- format(as.numeric(rownames(results.matrix)),digits=2)
ytick.pos <- seq(0,1,length.out = length(ytick.label))

# Plot
# The filled.contour() actoally draws 2 plots. 
# First, it calls ".filled.contour()" to draw the main plot, 
# then it draws the legend ("key").
# We can pass axis() and title() functions to the main plot and legend as follows:

(results.matrix, 
               axes = FALSE, # remove box around plot, clear axes from plot and legend
               levels = breaks, # levels of R_0 defining contours
               col = colors, # vector of colors (n = 1 - breaks)
               
               # Add custom axes, axis labels, and plot title
               plot.axes = { # surround R code to pass to main plot in brackets
                 axis(1,  # x axis
                      at = xtick.pos, labels = xtick.label, # tick positions and labels
                      lwd = 0, lwd.ticks = 1 # show tick marks but hide axis line
                      )
                 axis(2,  # y axis
                      at = ytick.pos, labels = ytick.label, 
                      lwd = 0, lwd.ticks = 1
                      )
                 title(main = latex2exp::TeX(sprintf(r'($\mu = %f$, $\beta = %f$)', mu, beta)), 
                       xlab = latex2exp::TeX(r'($\\sigma$)'),
                       ylab = latex2exp::TeX(r'($\\gamma$)'))
               },
               
               # Draw legend axes, labels, and legend title
               key.axes = { # surround R code to pass to legend (the "key") in brackets
                 axis(4, # right side
                      at=format(breaks, digits=2)) # place ticks marks at breaks
                 axis(2, # extra left side axis showing default ticks (i.e. "pretty") 
                      cex.axis = .65) # 
                 title(main=latex2exp::TeX(r'($R_0$)'))
               }

               )

```

```{r, echo=FALSE, eval=FALSE}
# 2. Add contours

# Plot
# Add contours by passing the contours() function to the plot.axes argument.
filled.contour(results.matrix, 
               axes = FALSE, 
               levels = breaks, 
               col = colors, 
               
               # Add custom axes, axis labels, and plot title
               plot.axes = {
                 axis(1,  
                      at = xtick.pos, labels = xtick.label, 
                      lwd = 0, lwd.ticks = 1 
                      )
                 axis(2,  # y axis
                      at = ytick.pos, labels = ytick.label, 
                      lwd = 0, lwd.ticks = 1
                      )
                 title(main = latex2exp::TeX(sprintf(r'($\mu = %f$, $\beta = %f$)', mu, beta)), 
                       xlab = latex2exp::TeX(r'($\\sigma$)'),
                       ylab = latex2exp::TeX(r'($\\gamma$)'))
                 
                 # Add contours
                 contour(results.matrix, add = TRUE, lwd = 1, 
                         levels = breaks,
                         labels = format(breaks, digits = 2))

               },
               
               # Draw legend axes, labels, and legend title
               key.axes = {
                 axis(4, # right side
                      at=format(breaks, digits=2)) # place ticks marks at breaks
                 axis(2, # extra left side axis showing default ticks (i.e. "pretty") 
                      cex.axis = .65) # 
                 title(main=latex2exp::TeX(r'($R_0$)'))
               }

               )

```

\pagebreak
## Exercise 4b

Consider models with more complex parameter spaces.

$\mathcal{R}_0$ for a model with both **environmental and direct transmission**
can be estimated as 
$$\mathcal{R}_0 = \frac{\beta}{(\mu + \gamma)} + \frac{\beta_v\omega}{\rho(\mu+\gamma)},$$

$\mathcal{R}_0$ for a **simple COVID model with asymptomatic transmission** can be estimated as 

$$\mathcal{R}_0 = \frac{\kappa \beta}{\gamma_A}+\frac{(1-\kappa)\beta}{\gamma}$$
How could you use the level plot approach to explore $\mathcal{R}_0$ across these parameter spaces? 

Can you imagine alternate ways to visualize $\mathcal{R}_0$ across these parameter spaces? Sketch your ideas on paper.


